Some examples in extended logic counting classes

### 2017-01-10 LICS/preliminaries.tex
sc

We assume that the reader is familiar with the semantics of $\so$, so we only introduce here some notation that will be used in this paper. 
%To define the semantics of $\so$, we need to introduce some terminology. 
Given a signature $\R$ and an $\R$-structure $\A$ with domain $A$, a first-order assignment $v$ for $\A$ is a total function from $\fv$ to $A$, while a second-order assignment $V$ for $\A$ is a total function with domain $\sv$ that maps each $X \in \sv$ to a subset of $A^{\arity(X)}$. Moreover, given a first-order assignment $v$ for $\A$, $x \in \fv$ and $a \in A$, we denote by $v[a/x]$ a first-order assignment such that $v[a/x](x) = a$ and $v[a/x](y) = v(y)$ for every $y \in \fv$ distinct from $x$. Similarly, given a second-order assignment $V$ for $\A$, $X \in \sv$ and $B  \subseteq A^{\arity(X)}$, we denote by $V[B/X]$ a second-order assignment such that $V[B/X](X) = B$ and $V[B/X](Y) = V(Y)$ for every $Y \in \sv$ distinct from $X$. We use notation $(\A, v, V) \models \varphi$ to indicate that structure $\A$ satisfies $\varphi$ under $v$ and $V$. In particular, we have that $(\A, v, V) \models \top$.
\martin{En vez de decir "for every $y\in \fv$ distinct from $x$" podría ser "for every $y\in\fv\setminus\{x\}$", y lo mismo con $Y\in\sv$.}

In this paper we consider several fragments of $\so$, in particular first-order logic ($\fo$), least fixed point logic (LFP) and partial fixed point logic (PFP) \cite{L04}. Moreover, for every $i \in \N$, we consider the fragment $\Sigma_i$ (resp., $\Pi_i$) of $\fo$, which is the set of $\fo$-formulae of the form 
$\exists \bar x_1 \forall \bar x_2 \cdots \exists \bar x_{i-1} \forall \bar x_{i} \, \psi$ (resp., 
$\forall \bar x_1 \exists \bar x_2 \cdots \forall \bar x_{i-1} \exists \bar x_{i} \, \psi$) if $i$ is even, and of the form
$\exists \bar x_1 \forall \bar x_2 \cdots \forall \bar x_{i-1} \exists \bar x_{i} \, \psi$ (resp., 
$\forall \bar x_1 \exists \bar x_2 \cdots \exists \bar x_{i-1} \forall \bar x_{i} \, \psi$) if $i$ is odd, where $\psi$ is a quantifier-free formula. Finally, we say that a fragment $\L_1$ is contained in a fragment $\L_2$, denoted by $\L_1 \subseteq \L_2$, if for every formula $\varphi$ in $\L_1$, there exists a formula $\psi$ in $\L_2$ such that $\varphi$ is logically equivalent to $\psi$.

@@

We consider the following counting complexity classes in this paper. 
%We define the following function complexity classes: 
$\fp$ is the class of functions computable in polynomial time, while $\fpspace$ is the class of functions computable in polynomial space. Moreover, $\nfpspace$ is the class of functions computable in polynomial space with output length bounded by a polynomial. Given a nondeterministic Turing Machine (NTM) $M$, let $\tma_M(x)$ be the number of accepting runs of $M$ with input $x$. Then $\shp$ is the class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tma_M(x)$ for every input $x$, while $\shl$ is the class of functions $f$ for which there exists a logarithmic-space NTM $M$ such that $f(x) = \tma_M(x)$ for every input $x$.  Given an NTM $M$ with output tape, let $\tmo_M(x)$ be the number of distinct outputs of $M$ with input $x$ (notice that  $M$ produces an output if it halts in an accepting state). Then $\spp$ is the  class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tmo_M(x)$ for every input $x$. Notice that $\shp \subseteq \spp$, and this inclusion is believed to be strict.

>>>>

We assume that the reader is familiar with the semantics of $\so$, so we only introduce here some notation that will be used in this paper. 
%To define the semantics of $\so$, we need to introduce some terminology. 
Given a signature $\R$ and an $\R$-structure $\A$ with domain $A$, a first-order assignment $v$ for $\A$ is a total function from $\fv$ to $A$, while a second-order assignment $V$ for $\A$ is a total function with domain $\sv$ that maps each $X \in \sv$ to a subset of $A^{\arity(X)}$. Moreover, given a first-order assignment $v$ for $\A$, $x \in \fv$ and $a \in A$, we denote by $v[a/x]$ a first-order assignment such that $v[a/x](x) = a$ and $v[a/x](y) = v(y)$ for every $y \in \fv$ distinct from $x$. Similarly, given a second-order assignment $V$ for $\A$, $X \in \sv$ and $B  \subseteq A^{\arity(X)}$, we denote by $V[B/X]$ a second-order assignment such that $V[B/X](X) = B$ and $V[B/X](Y) = V(Y)$ for every $Y \in \sv$ distinct from $X$. We use notation $(\A, v, V) \models \varphi$ to indicate that structure $\A$ satisfies $\varphi$ under $v$ and $V$. In particular, we have that $(\A, v, V) \models \top$.
%\martin{En vez de decir "for every $y\in \fv$ distinct from $x$" podría ser "for every $y\in\fv\setminus\{x\}$", y lo mismo con $Y\in\sv$.}

In this paper, we consider several fragments of $\so$, in particular first-order logic ($\fo$), least fixed point logic (LFP) and partial fixed point logic (PFP) \cite{L04}. Moreover, for every $i \in \N$, we consider the fragment $\Sigma_i$ (resp., $\Pi_i$) of $\fo$, which is the set of $\fo$-formulae of the form 
$\exists \bar x_1 \forall \bar x_2 \cdots \exists \bar x_{i-1} \forall \bar x_{i} \, \psi$ (resp., 
$\forall \bar x_1 \exists \bar x_2 \cdots \forall \bar x_{i-1} \exists \bar x_{i} \, \psi$) if $i$ is even, and of the form
$\exists \bar x_1 \forall \bar x_2 \cdots \forall \bar x_{i-1} \exists \bar x_{i} \, \psi$ (resp., 
$\forall \bar x_1 \exists \bar x_2 \cdots \exists \bar x_{i-1} \forall \bar x_{i} \, \psi$) if $i$ is odd, where $\psi$ is a quantifier-free formula. Finally, we say that a fragment $\L_1$ is contained in a fragment $\L_2$, denoted by $\L_1 \subseteq \L_2$, if for every formula $\varphi$ in $\L_1$, there exists a formula $\psi$ in $\L_2$ such that $\varphi$ is logically equivalent to $\psi$.  Besides, we say that $\L_1$ is properly contained in $\L_2$, denoted by $\L_1 \subsetneq \L_2$, if $\L_1 \subseteq \L_2$ and $\L_2 \not\subseteq \L_1$.

@@

We consider the following counting complexity classes in this paper. 
%We define the following function complexity classes: 
$\fp$ is the class of functions $f : \Sigma^* \to \N$ computable in polynomial time, while $\fpspace$ is the class of functions $f : \Sigma^* \to \N$ computable in polynomial space. 
%Moreover, $\nfpspace$ is the class of functions computable in polynomial space with output length bounded by a polynomial. 
Given a nondeterministic Turing Machine (NTM) $M$, let $\tma_M(x)$ be the number of accepting runs of $M$ with input $x$. Then $\shp$ is the class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tma_M(x)$ for every input $x$, while $\shl$ is the class of functions $f$ for which there exists a logarithmic-space NTM $M$ such that $f(x) = \tma_M(x)$ for every input $x$.  Given an NTM $M$ with output tape, let $\tmo_M(x)$ be the number of distinct outputs of $M$ with input $x$ (notice that  $M$ produces an output if it halts in an accepting state). Then $\spp$ is the  class of functions $f$ for which there exists a polynomial-time NTM $M$ such that $f(x) = \tmo_M(x)$ for every input $x$. Notice that $\shp \subseteq \spp$, and this inclusion is believed to be strict.
